---
title: "AS Carbon Audit 2024-2025"
author: "Amy Zhao, Director of Research and Data Analytics <br \n> Ji Moon, Technical Projects Coordinator <br \n> Dhruv S. Kannan, Associate Vice President <br \n> Office of Environmental Justice Affairs"
date: today
format:
  revealjs:
    theme: [simple, theme.scss]
    logo: www/eja_logo.png
    slide-number: true
    resources:
      - www/eja_logo.png
    incremental: false
    center: true
    height: 800
    fig-align: center
    echo: false
    transition: fade
    auto-stretch: true
server: shiny
---

```{r}
#| context: setup
#| include: false
library(shiny)
library(DBI)
library(RPostgres)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(plotly)

db_connect <- function() {
  dbConnect(
    Postgres(),
    host     = Sys.getenv("SUPABASE_HOST"),
    port     = as.integer(Sys.getenv("SUPABASE_PORT", "5432")),
    dbname   = Sys.getenv("SUPABASE_DB", "postgres"),
    user     = Sys.getenv("SUPABASE_USER"),
    password = Sys.getenv("SUPABASE_PASSWORD"),
    sslmode  = "require"
  )
}

# Real per-transaction emissions live in `climatiq`.
# Every schema has a `category` column populated with one of the 7 canonical
# All queries exclude rows where 'climatiq mapping' = 'exclude'.

EMISSIONS_SQL <-
  'select "Office" as office, "climatiq mapping", category, "Cost" as cost, co2e
     from climatiq.merged
    where co2e is not null
      and "climatiq mapping" != \'exclude\''

PALETTE <- c(
  "Travel & Transportation" = "#d62728",
  "Professional Services"   = "#1f77b4",
  "Events"                  = "#ff7f0e",
  "Equipment & Rentals"     = "#2ca02c",
  "Food & Catering"         = "#9467bd",
  "Goods & Merchandise"     = "#8c564b",
  "Other"                   = "#7f7f7f"
)
```

```{r}
#| context: server
data_pull <- eventReactive(input$refresh, {
  required <- c("SUPABASE_HOST", "SUPABASE_USER", "SUPABASE_PASSWORD")
  missing  <- required[Sys.getenv(required) == ""]
  if (length(missing)) {
    stop(paste(
      "Missing env vars:", paste(missing, collapse = ", ")
    ))
  }

  con <- tryCatch(db_connect(), error = function(e) {
    stop(paste("DB connect failed:", conditionMessage(e)))
  })
  on.exit(dbDisconnect(con), add = TRUE)

  emissions <- tryCatch(dbGetQuery(con, EMISSIONS_SQL), error = function(e) {
    stop(paste("Query failed:", conditionMessage(e)))
  })

  agg <- emissions |>
    group_by(office, category) |>
    summarise(co2e = sum(co2e, na.rm = TRUE), n = n(), .groups = "drop")

  list(
    emissions  = emissions,
    agg        = agg,
    fetched_at = Sys.time()
  )
}, ignoreNULL = FALSE)

office_order <- reactive({
  data_pull()$agg |>
    group_by(office) |> summarise(co2e = sum(co2e), .groups = "drop") |>
    arrange(desc(co2e)) |> pull(office)
})

category_order <- reactive({
  data_pull()$agg |>
    group_by(category) |> summarise(co2e = sum(co2e), .groups = "drop") |>
    arrange(desc(co2e)) |> pull(category)
})

output$fetched_at <- renderText({
  paste0("Last refreshed: ", format(data_pull()$fetched_at, "%Y-%m-%d %H:%M:%S"))
})

output$tx_count   <- renderText(format(nrow(data_pull()$emissions), big.mark = ","))
output$total_co2  <- renderText(format(round(sum(data_pull()$emissions$co2e)), big.mark = ","))
output$total_co2_ctx <- renderText(format(round(sum(data_pull()$emissions$co2e)), big.mark = ","))
output$total_cost <- renderText(paste0("$", format(round(sum(data_pull()$emissions$cost, na.rm = TRUE)), big.mark = ",")))
output$mean_co2   <- renderText(format(round(mean(data_pull()$emissions$co2e, na.rm = TRUE), 1), big.mark = ","))
output$p90_co2    <- renderText(format(round(quantile(data_pull()$emissions$co2e, 0.9, na.rm = TRUE), 1), big.mark = ","))
output$sd_co2     <- renderText(format(round(sd(data_pull()$emissions$co2e, na.rm = TRUE), 1), big.mark = ","))

output$boxplot <- renderPlot({
  positive <- data_pull()$emissions |>
    filter(co2e > 0) |>
    mutate(office = factor(office, levels = rev(office_order())))

  ggplot(positive, aes(x = co2e, y = office)) +
    geom_boxplot(fill = "#2b8a3e", colour = "#1b4332", alpha = 0.85,
                 width = 0.65,
                 outlier.colour = "#888888", outlier.alpha = 0.4,
                 outlier.size = 1) +
    scale_x_log10(labels = label_comma()) +
    labs(x = "Emissions per transaction (kg CO2e, log scale)",
         y = NULL,
         title = "Per-transaction emissions by office — shared log x-axis") +
    theme_minimal(base_size = 12) +
    theme(panel.grid.major.y = element_blank(),
          panel.grid.minor.x = element_line(linetype = "dotted"),
          plot.title.position = "plot")
})

output$share_by_office <- renderPlot({
  oo <- office_order()
  totals <- data_pull()$emissions |>
    filter(co2e > 0, cost > 0) |>
    group_by(office) |>
    summarise(spend = sum(cost, na.rm = TRUE),
              co2e  = sum(co2e, na.rm = TRUE), .groups = "drop") |>
    mutate(`Share of spend`    = spend / sum(spend),
           `Share of emissions`= co2e  / sum(co2e)) |>
    select(office, `Share of spend`, `Share of emissions`) |>
    pivot_longer(-office, names_to = "metric", values_to = "share") |>
    mutate(office = factor(office, levels = rev(oo)),
           metric = factor(metric, levels = c("Share of spend", "Share of emissions")))

  ggplot(totals, aes(x = share, y = office, fill = metric)) +
    geom_col(position = position_dodge(width = 0.75), width = 0.7) +
    scale_x_continuous(labels = percent_format(accuracy = 1)) +
    scale_fill_manual(values = c("Share of spend" = "#8ab17d",
                                 "Share of emissions" = "#bc4749")) +
    labs(x = NULL, y = NULL, fill = NULL,
         title = "Share of spend vs. share of emissions, by office",
         subtitle = "Offices where the red bar exceeds the green bar punch above their spending weight on carbon") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "top",
          plot.title.position = "plot",
          panel.grid.major.y = element_blank())
})

output$intensity_by_category <- renderPlot({
  int_df <- data_pull()$emissions |>
    filter(co2e > 0, cost > 0) |>
    group_by(category) |>
    summarise(spend = sum(cost, na.rm = TRUE),
              co2e  = sum(co2e, na.rm = TRUE), .groups = "drop") |>
    mutate(intensity = co2e / spend) |>
    arrange(intensity) |>
    mutate(category = factor(category, levels = category))

  overall <- sum(int_df$co2e) / sum(int_df$spend)

  ggplot(int_df, aes(x = intensity, y = category, fill = category)) +
    geom_col(width = 0.7) +
    geom_vline(xintercept = overall, linetype = "dashed", colour = "#333333") +
    annotate("text", x = overall, y = 0.7,
             label = sprintf("overall avg: %.2f kg/$", overall),
             hjust = -0.05, vjust = 0, size = 3.5, colour = "#333333") +
    scale_fill_manual(values = PALETTE, guide = "none") +
    scale_x_continuous(labels = label_comma()) +
    labs(x = "kg CO2e per dollar spent", y = NULL,
         title = "Carbon intensity by expenditure category",
         subtitle = "How much CO2e each dollar buys, summed across all offices") +
    theme_minimal(base_size = 12) +
    theme(plot.title.position = "plot",
          panel.grid.major.y = element_blank())
})

output$stacked <- renderPlot({
  oo <- office_order(); co <- category_order()
  stack_df <- data_pull()$agg |>
    mutate(office   = factor(office,   levels = oo),
           category = factor(category, levels = rev(co)))

  ggplot(stack_df, aes(x = office, y = co2e, fill = category)) +
    geom_col(colour = "white", linewidth = 0.3) +
    scale_fill_manual(values = PALETTE, breaks = co) +
    scale_y_continuous(labels = label_comma()) +
    labs(x = NULL, y = "Total emissions (kg CO2e)",
         fill = NULL,
         title = "Total emissions by office, split by expenditure category") +
    theme_minimal(base_size = 12) +
    theme(axis.text.x = element_text(angle = 35, hjust = 1),
          panel.grid.major.x = element_blank(),
          panel.grid.minor.y = element_blank(),
          legend.position = "top",
          plot.title.position = "plot")
})

output$sankey <- renderPlotly({
  oo <- office_order(); co <- category_order()
  nodes       <- c(oo, co)
  node_index  <- setNames(seq_along(nodes) - 1L, nodes)
  cat_cols    <- unname(PALETTE[co]); cat_cols[is.na(cat_cols)] <- "#cccccc"
  node_colors <- c(rep("#1b4332", length(oo)), cat_cols)

  hex_to_rgba <- function(h, a = 0.45) {
    h <- sub("^#", "", h)
    sprintf("rgba(%d,%d,%d,%.2f)",
            strtoi(substr(h, 1, 2), 16),
            strtoi(substr(h, 3, 4), 16),
            strtoi(substr(h, 5, 6), 16), a)
  }

  links <- data_pull()$agg |>
    filter(co2e > 0) |>
    mutate(source = node_index[office],
           target = node_index[category],
           color  = vapply(PALETTE[category], hex_to_rgba, character(1), a = 0.45)) |>
    mutate(color = ifelse(is.na(color), hex_to_rgba("#cccccc"), color))

  plot_ly(
    type = "sankey", arrangement = "snap",
    node = list(label = nodes, color = node_colors,
                pad = 14, thickness = 14,
                line = list(color = "white", width = 0.5)),
    link = list(source = links$source, target = links$target,
                value  = links$co2e,   color  = links$color,
                hovertemplate = "%{source.label} → %{target.label}<br>%{value:,.0f} kg CO2e<extra></extra>")
  ) |>
    layout(title = list(text = "Emissions flow: office → expenditure category (kg CO2e)", x = 0.01),
           font  = list(size = 11),
           margin = list(l = 10, r = 10, t = 50, b = 10),
           height = 640)
})
```

## Live Data {.smaller}

```{r}
actionButton("refresh", "Refresh from Supabase", icon = icon("rotate"))
textOutput("fetched_at", inline = TRUE)
```

------------------------------------------------------------------------

## Why are we Doing This?

-   **Sustainability:** making sure the needs of the present do not outweigh the needs of the future
-   Contextualize the planet-warming potential of the Association
-   Make sure our money serves a livable climate
-   Get in compliance with campus climate goals:
    -   Revising Climate Action Plan for first time in a decade
-   Goal: **Become first carbon-neutral student government in the country**
    -   First-of-its-kind initiative among UC student governments

------------------------------------------------------------------------

## What this Audit is Not

- An inquisition
- A forum to air grievances
- Marching orders
  - Solutions will be implemented in a collaborative, deliberative manner

------------------------------------------------------------------------

## How we did it

- All financial details implemented in Supabase, a data management system  
- Use the **Climatiq API**
-   Two-step process:
    1.  Feed purchase description into system, which matches it to an **emissions factor** (i.e., beef jerky → `consumer_goods-type_snack_foods`)
    2.  System calculates $\text{CO}_2$ equivalent per purchase
-   Use Python/R for visualizations and statistical analysis, with live data fetch from Supabase

------------------------------------------------------------------------

## Overview

-   **22 offices** across Associated Students, FY 2024-2025
-   **Transactions:** `r textOutput("tx_count", inline = TRUE)`
-   **Total kg CO~2~e:** `r textOutput("total_co2", inline = TRUE)` (spend-based Scope I)
-   Three views in this deck:
    1.  Distribution of per-transaction emissions by office (boxplot)
    2.  Total emissions by office, split by expenditure category (stacked bar)
    3.  Flow from offices to expenditure categories (Sankey)

------------------------------------------------------------------------

## Headline Numbers

::::::::::: stats-row
:::::: stats-col
::: stat
[`r textOutput("tx_count", inline = TRUE)`]{.stat-number} [transactions analyzed]{.stat-label}
:::

::: stat
[`r textOutput("total_co2", inline = TRUE)`]{.stat-number} [kg CO₂e total emissions]{.stat-label}
:::

::: stat
[`r textOutput("total_cost", inline = TRUE)`]{.stat-number} [total carbon spend]{.stat-label}
:::
::::::

:::::: stats-col
::: stat
[`r textOutput("mean_co2", inline = TRUE)`]{.stat-number} [mean kg CO₂e per transaction]{.stat-label}
:::

::: stat
[`r textOutput("p90_co2", inline = TRUE)`]{.stat-number} [90th percentile kg CO₂e per transaction]{.stat-label}
:::

::: stat
[`r textOutput("sd_co2", inline = TRUE)`]{.stat-number} [SD kg CO₂e per transaction]{.stat-label}
:::
::::::
:::::::::::

------------------------------------------------------------------------

## Contextualizing the Carbon

::: hero-stat
[`r textOutput("total_co2_ctx", inline = TRUE)`]{.stat-number}

[kg CO₂e total emissions]{.stat-label}
:::

-   Equivalent to driving a gas-powered car roughly **730,000 miles**
-   Same as the annual electricity use of around **40 U.S. homes**
-   Would take about **286 acres of U.S. forests** a full year to sequester
-   417 round trips from SD to NY
-   Roughly **23,000,000 smartphones charged**
    -   Source: EPA [Greenhouse Gas Equivalencies Calculator](https://www.epa.gov/energy/greenhouse-gas-equivalencies-calculator#results)

------------------------------------------------------------------------

## Total Emissions by Office --- Stacked by Category {.smaller}

```{r}
plotOutput("stacked", height = "520px")
```

::: aside
Stacks ordered (bottom → top) by global category total.
:::

------------------------------------------------------------------------

## Flow: Office → Category (Sankey)

```{r}
plotlyOutput("sankey", height = "640px")
```

---

## Emissions by Office --- Distribution {.smaller}

```{r}
plotOutput("boxplot", height = "520px")
```

::: aside
Offices ordered by total CO~2~e (highest at top). Zero / negative values excluded from log axis.
:::

------------------------------------------------------------------------

## Spend Share vs. Emissions Share, by Office {.smaller}

```{r}
plotOutput("share_by_office", height = "560px")
```

::: aside
Each office's share of total spend (green) vs. share of total emissions (red). Red > green means the office is more carbon-intensive than its spending alone would suggest.
:::

------------------------------------------------------------------------

## Carbon Intensity by Category {.smaller}

```{r}
plotOutput("intensity_by_category", height = "520px")
```

::: aside
kg CO~2~e per dollar spent, category totals. Dashed line is the overall average across all transactions.
:::

------------------------------------------------------------------------

## Takeaways

-   **Travel & Transportation** and **Professional Services** dominate totals at VP External and Transportation & Transformation.
-   **ASCE** has the largest and most dispersed per-transaction distribution --- many small Events and Equipment rentals plus a long outlier tail.
-   **Equipment & Rentals** is the largest single category by transaction count.
-   Negative / zero `co2e` rows (credits, reversals, unmapped) were excluded from the log-scale view and shown as Uncategorized in totals.

------------------------------------------------------------------------

## Potential Solutions

1. Reduce Amazon spending
2. Identify more sustainable travel options
  - Book flights well in advance
  - Utilize carbon forecasts online to identify better options:
    - Earlier flights
    - Newer aircraft
    - Fuel-efficient vehicles

------------------------------------------------------------------------

## Next Steps

- EJA + SSC -- implement **Waste Reduction Protocols**
  - Identify sustainable alternatives to Amazon purchases
  - Target: **end of the school year**
- EJA + Senate Special Committee on Environmental Justice - identify opportunities for carbon offsets, reinvesting in green initiatives
  - Revive SSC's Green Initiative Fund to fund student-powered sustainability projects, use A.S. institutional position and funding to implement as necessary
- Conversations to come! 

------------------------------------------------------------------------

## Thank you!

Questions? Reach out to Dhruv on Slack. This presentation will be shared for review and will be pushed online when possible.

------------------------------------------------------------------------