---
title: "AS Carbon Audit 2024-2025"
subtitle: "The Carbon Impact of the Association"
author: "Amy Zhao, Director of Research and Data Analytics <br \n> Ji Moon, Technical Projects Coordinator \n Office of Environmental Justice Affairs"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    incremental: false
    width: 1280
    height: 800
    fig-align: center
    echo: false
    transition: fade
    auto-stretch: false
server: shiny
---

```{r}
#| context: setup
#| include: false
library(shiny)
library(DBI)
library(RPostgres)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(plotly)

db_connect <- function() {
  dbConnect(
    Postgres(),
    host     = Sys.getenv("SUPABASE_HOST"),
    port     = as.integer(Sys.getenv("SUPABASE_PORT", "5432")),
    dbname   = Sys.getenv("SUPABASE_DB", "postgres"),
    user     = Sys.getenv("SUPABASE_USER"),
    password = Sys.getenv("SUPABASE_PASSWORD"),
    sslmode  = "require"
  )
}

# Real per-transaction emissions live in `<office>_2024_2025.carbon_emission`.
# Every schema has a `category` column populated with one of the 7 canonical
# categories. All queries exclude rows flagged `remove = true`.
OFFICE_SCHEMAS <- c(
  "ASCE"                           = "asce_2024_2025",
  "ASOSA"                          = "asosa_2024_2025",
  "CCMO"                           = "ccmo_2024_2025",
  "CFO"                            = "cfo_2024_2025",
  "College Affairs"                = "college_affairs_2024_2025",
  "CPO"                            = "cpo_2024_2025",
  "EDI"                            = "edi_2024_2025",
  "Enterprises & Services"         = "enterprises_and_services_2024_2025",
  "Environmental Justice Affairs"  = "environmental_justice_affairs_2024_2025",
  "EVP"                            = "evp_2024_2025",
  "Food & Housing"                 = "food_and_housing_2024_2025",
  "Health & Wellbeing"             = "health_and_wellbeing_2024_2025",
  "Local Affairs"                  = "local_affairs_2024_2025",
  "Mandate Reserves"               = "mandate_reserves_2024_2025",
  "Office of the President"        = "office_of_the_president_2024_2025",
  "SSC"                            = "ssc_2024_2025",
  "Student Advocacy"               = "student_advocacy_2024_2025",
  "Student Orgs"                   = "student_orgs_2024_2025",
  "Transportation & Transformation"= "transportation_and_transformation_2024_2025",
  "TTV"                            = "ttv_2024_2025",
  "VPAA"                           = "vpaa_2024_2025",
  "VP External"                    = "vp_external_2024_2025"
)

EMISSIONS_SQL <- paste(
  sprintf(
    "select '%s'::text as office, category, cost, co2e
       from %s.carbon_emission
      where coalesce(remove, false) = false
        and co2e is not null",
    names(OFFICE_SCHEMAS), unname(OFFICE_SCHEMAS)
  ),
  collapse = "\nunion all\n"
)

PALETTE <- c(
  "Travel & Transportation" = "#d62728",
  "Professional Services"   = "#1f77b4",
  "Events"                  = "#ff7f0e",
  "Equipment & Rentals"     = "#2ca02c",
  "Food & Catering"         = "#9467bd",
  "Goods & Merchandise"     = "#8c564b",
  "Other"                   = "#7f7f7f"
)
```

```{r}
#| context: server
data_pull <- eventReactive(input$refresh, {
  con <- db_connect()
  on.exit(dbDisconnect(con), add = TRUE)

  emissions <- dbGetQuery(con, EMISSIONS_SQL)

  agg <- emissions |>
    group_by(office, category) |>
    summarise(co2e = sum(co2e, na.rm = TRUE), n = n(), .groups = "drop")

  list(
    emissions  = emissions,
    agg        = agg,
    fetched_at = Sys.time()
  )
}, ignoreNULL = FALSE)

office_order <- reactive({
  data_pull()$agg |>
    group_by(office) |> summarise(co2e = sum(co2e), .groups = "drop") |>
    arrange(desc(co2e)) |> pull(office)
})

category_order <- reactive({
  data_pull()$agg |>
    group_by(category) |> summarise(co2e = sum(co2e), .groups = "drop") |>
    arrange(desc(co2e)) |> pull(category)
})

output$fetched_at <- renderText({
  paste0("Last refreshed: ", format(data_pull()$fetched_at, "%Y-%m-%d %H:%M:%S"))
})

output$tx_count  <- renderText(format(nrow(data_pull()$emissions), big.mark = ","))
output$total_co2 <- renderText(format(round(sum(data_pull()$emissions$co2e)), big.mark = ","))

output$boxplot <- renderPlot({
  positive <- data_pull()$emissions |>
    filter(co2e > 0) |>
    mutate(office = factor(office, levels = rev(office_order())))

  ggplot(positive, aes(x = co2e, y = office)) +
    geom_boxplot(fill = "#2b8a3e", colour = "#1b4332", alpha = 0.85,
                 width = 0.65,
                 outlier.colour = "#888888", outlier.alpha = 0.4,
                 outlier.size = 1) +
    scale_x_log10(labels = label_comma()) +
    labs(x = "Emissions per transaction (kg CO2e, log scale)",
         y = NULL,
         title = "Per-transaction emissions by office — shared log x-axis") +
    theme_minimal(base_size = 12) +
    theme(panel.grid.major.y = element_blank(),
          panel.grid.minor.x = element_line(linetype = "dotted"),
          plot.title.position = "plot")
})

output$stacked <- renderPlot({
  oo <- office_order(); co <- category_order()
  stack_df <- data_pull()$agg |>
    mutate(office   = factor(office,   levels = oo),
           category = factor(category, levels = rev(co)))

  ggplot(stack_df, aes(x = office, y = co2e, fill = category)) +
    geom_col(colour = "white", linewidth = 0.3) +
    scale_fill_manual(values = PALETTE, breaks = co) +
    scale_y_continuous(labels = label_comma()) +
    labs(x = NULL, y = "Total emissions (kg CO2e)",
         fill = NULL,
         title = "Total emissions by office, split by expenditure category") +
    theme_minimal(base_size = 12) +
    theme(axis.text.x = element_text(angle = 35, hjust = 1),
          panel.grid.major.x = element_blank(),
          panel.grid.minor.y = element_blank(),
          legend.position = "top",
          plot.title.position = "plot")
})

output$sankey <- renderPlotly({
  oo <- office_order(); co <- category_order()
  nodes       <- c(oo, co)
  node_index  <- setNames(seq_along(nodes) - 1L, nodes)
  cat_cols    <- unname(PALETTE[co]); cat_cols[is.na(cat_cols)] <- "#cccccc"
  node_colors <- c(rep("#1b4332", length(oo)), cat_cols)

  hex_to_rgba <- function(h, a = 0.45) {
    h <- sub("^#", "", h)
    sprintf("rgba(%d,%d,%d,%.2f)",
            strtoi(substr(h, 1, 2), 16),
            strtoi(substr(h, 3, 4), 16),
            strtoi(substr(h, 5, 6), 16), a)
  }

  links <- data_pull()$agg |>
    filter(co2e > 0) |>
    mutate(source = node_index[office],
           target = node_index[category],
           color  = vapply(PALETTE[category], hex_to_rgba, character(1), a = 0.45)) |>
    mutate(color = ifelse(is.na(color), hex_to_rgba("#cccccc"), color))

  plot_ly(
    type = "sankey", arrangement = "snap",
    node = list(label = nodes, color = node_colors,
                pad = 14, thickness = 14,
                line = list(color = "white", width = 0.5)),
    link = list(source = links$source, target = links$target,
                value  = links$co2e,   color  = links$color,
                hovertemplate = "%{source.label} → %{target.label}<br>%{value:,.0f} kg CO2e<extra></extra>")
  ) |>
    layout(title = list(text = "Emissions flow: office → expenditure category (kg CO2e)", x = 0.01),
           font  = list(size = 11),
           margin = list(l = 10, r = 10, t = 50, b = 10),
           height = 640)
})
```

## Live data {.smaller}

```{r}
actionButton("refresh", "Refresh from Supabase", icon = icon("rotate"))
textOutput("fetched_at", inline = TRUE)
```

------------------------------------------------------------------------

## Why are we doing this?

-   **Sustainability:** making sure the needs of the present do not outweigh the needs of the future
-   Contextualize the planet-warming potential of the Association
-   Make sure our money serves a livable climate
-   Get in compliance with campus climate goals:
    -   Revising Climate Action Plan for first time in a decade
-   Goal: **Become first carbon-neutral student government in the country**
    -   First-of-its-kind initiative among UC student governments

---

## How we did it

-   Use the **Climatiq API**
-   Two-step process:
    1.  Feed purchase description into system, which matches it to an **emissions factor** (i.e., beef jerky → `consumer_goods-type_snack_foods`)
    2.  System calculates $\text{CO}_2$ equivalent per purchase
-   Use Python/R for visualizations and statistical analysis

---

## Overview

-   **22 offices** across Associated Students, FY 2024-2025
-   **Transactions:** `r textOutput("tx_count", inline = TRUE)`
-   **Total kg CO~2~e:** `r textOutput("total_co2", inline = TRUE)` (spend-based Scope I)
-   Three views in this deck:
    1.  Distribution of per-transaction emissions by office (boxplot)
    2.  Total emissions by office, split by expenditure category (stacked bar)
    3.  Flow from offices to expenditure categories (Sankey)

------------------------------------------------------------------------

## Emissions by office --- distribution {.smaller}

```{r}
plotOutput("boxplot", height = "520px")
```

::: aside
Offices ordered by total CO~2~e (highest at top). Zero / negative values excluded from log axis.
:::

------------------------------------------------------------------------

## Total emissions by office --- stacked by category {.smaller}

```{r}
plotOutput("stacked", height = "520px")
```

::: aside
Stacks ordered (bottom → top) by global category total.
:::

------------------------------------------------------------------------

## Flow: office → category (Sankey)

```{r}
plotlyOutput("sankey", height = "640px")
```

------------------------------------------------------------------------

## Takeaways

-   **Travel & Transportation** and **Professional Services** dominate totals at VP External and Transportation & Transformation.
-   **ASCE** has the largest and most dispersed per-transaction distribution --- many small Events and Equipment rentals plus a long outlier tail.
-   **Equipment & Rentals** is the largest single category by transaction count.
-   Negative / zero `co2e` rows (credits, reversals, unmapped) were excluded from the log-scale view and shown as Uncategorized in totals.