---
title: "AS Carbon Audit 2024-2025"
subtitle: "The Carbon Impact of the Association"
author: "Amy Zhao, Director of Research and Data Analytics; and Ji Moon, Technical Projects Coordinator, for the Office of Environmental Justice Affairs"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    incremental: false
    width: 1280
    height: 800
    embed-resources: true
    fig-align: center
    echo: false
    transition: fade
---

```{r}
#| label: setup
#| include: false
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(plotly)

emissions <- read_csv("data/emissions.csv", show_col_types = FALSE)
agg       <- read_csv("data/office_category.csv", show_col_types = FALSE)

office_totals <- agg |>
  group_by(office) |>
  summarise(co2e = sum(co2e), .groups = "drop") |>
  arrange(desc(co2e))
office_order <- office_totals$office

category_totals <- agg |>
  group_by(category) |>
  summarise(co2e = sum(co2e), .groups = "drop") |>
  arrange(desc(co2e))
category_order <- category_totals$category

PALETTE <- c(
  "Travel & Transportation" = "#d62728",
  "Professional Services"   = "#1f77b4",
  "Events"                  = "#ff7f0e",
  "Equipment & Rentals"     = "#2ca02c",
  "Food & Catering"         = "#9467bd",
  "Goods & Merchandise"     = "#8c564b",
  "Other"                   = "#7f7f7f",
  "Uncategorized"           = "#bcbd22"
)
```

## Why are we doing this?

-   **Sustainability:** making sure the needs of the present do not outweigh the needs of the future
-   Contextualize the planet-warming potential of the Association
-   Make sure our money serves a livable climate
-   Get in compliance with campus climate goals:
    -   Revising Climate Action Plan for first time in a decade
-   Goal: **Become first carbon-neutral student government in the country**
    -   First-of-its-kind initiative among UC student governments

## How we did it

-   Use the **Climatiq API**
-   Two-step process:
    1.  Feed purchase description into system, which matches it to an **emissions factor** (i.e., for every \$ spent on this item, how much $\text{CO}_2\text{e}$ is emitted?)
    2.  System calculates $\text{CO}_2$ equivalent per purchase
-   Use Python/R for visualizations and statistical analysis

------------------------------------------------------------------------

## Overview

-   **22 offices** across Associated Students, FY 2024-2025
-   **`r format(nrow(emissions), big.mark = ",")` transactions** with a computed emission factor
-   **`r format(round(sum(emissions$co2e)), big.mark = ",")` kg CO~2~e** total (spend-based Scope 3)
-   Three views in this deck:
    1.  Distribution of per-transaction emissions by office (boxplot)
    2.  Total emissions by office, split by expenditure category (stacked bar)
    3.  Flow from offices to expenditure categories (Sankey)

------------------------------------------------------------------------

## Emissions by office --- distribution {.smaller}

```{r}
#| fig-width: 12
#| fig-height: 6.5
positive <- emissions |>
  filter(co2e > 0) |>
  mutate(office = factor(office, levels = rev(office_order)))

ggplot(positive, aes(x = co2e, y = office)) +
  geom_boxplot(fill = "#2b8a3e", colour = "#1b4332", alpha = 0.85,
               width = 0.65,
               outlier.colour = "#888888", outlier.alpha = 0.4,
               outlier.size = 1) +
  scale_x_log10(labels = label_comma()) +
  labs(x = "Emissions per transaction (kg CO2e, log scale)",
       y = NULL,
       title = "Per-transaction emissions by office — shared log x-axis") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.x = element_line(linetype = "dotted"),
        plot.title.position = "plot")
```

::: aside
Offices ordered by total CO~2~e (highest at top). Zero / negative values excluded from log axis.
:::

------------------------------------------------------------------------

## Total emissions by office --- stacked by category {.smaller}

```{r}
#| fig-width: 12
#| fig-height: 6.5
stack_df <- agg |>
  mutate(office   = factor(office,   levels = office_order),
         category = factor(category, levels = rev(category_order)))

ggplot(stack_df, aes(x = office, y = co2e, fill = category)) +
  geom_col(colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = PALETTE, breaks = category_order) +
  scale_y_continuous(labels = label_comma()) +
  labs(x = NULL, y = "Total emissions (kg CO2e)",
       fill = NULL,
       title = "Total emissions by office, split by expenditure category") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position = "top",
        plot.title.position = "plot")
```

::: aside
Stacks ordered (bottom → top) by global category total.
:::

------------------------------------------------------------------------

## Flow: office → category (Sankey)

```{r}
#| fig-width: 12
#| fig-height: 7.2
nodes       <- c(office_order, category_order)
node_index  <- setNames(seq_along(nodes) - 1L, nodes)
office_col  <- "#1b4332"
cat_cols    <- unname(PALETTE[category_order])
cat_cols[is.na(cat_cols)] <- "#cccccc"
node_colors <- c(rep(office_col, length(office_order)), cat_cols)

hex_to_rgba <- function(h, a = 0.45) {
  h <- sub("^#", "", h)
  r <- strtoi(substr(h, 1, 2), 16)
  g <- strtoi(substr(h, 3, 4), 16)
  b <- strtoi(substr(h, 5, 6), 16)
  sprintf("rgba(%d,%d,%d,%.2f)", r, g, b, a)
}

links <- agg |>
  filter(co2e > 0) |>
  mutate(source = node_index[office],
         target = node_index[category],
         color  = vapply(PALETTE[category], hex_to_rgba, character(1), a = 0.45)) |>
  mutate(color = ifelse(is.na(color), hex_to_rgba("#cccccc"), color))

plot_ly(
  type = "sankey",
  arrangement = "snap",
  node = list(label = nodes, color = node_colors,
              pad = 14, thickness = 14,
              line = list(color = "white", width = 0.5)),
  link = list(source = links$source, target = links$target,
              value  = links$co2e,   color  = links$color,
              hovertemplate = "%{source.label} → %{target.label}<br>%{value:,.0f} kg CO2e<extra></extra>")
) |>
  layout(title = list(text = "Emissions flow: office → expenditure category (kg CO2e)", x = 0.01),
         font  = list(size = 11),
         margin = list(l = 10, r = 10, t = 50, b = 10),
         height = 640)
```

------------------------------------------------------------------------

## Takeaways

-   **Travel & Transportation** and **Professional Services** dominate totals at VP External and Transportation & Transformation.
-   **ASCE** has the largest and most dispersed per-transaction distribution --- many small Events and Equipment rentals plus a long outlier tail.
-   **Equipment & Rentals** is the largest single category by transaction count.
-   Negative / zero `co2e` rows (credits, reversals, unmapped) were excluded from the log-scale view and shown as Uncategorized in totals.